Exploring mathematical objects from custom-tailored mathematical universes

### 2020-01-19 paper-filmat.tex
typo

\begin{tabular}{lll}
  \toprule
  Statement & in $\Set$ & in $\Eff$ \\
  \midrule
  Any natural number is prime or not prime. & \cmark{} (trivially so) & \cmark \\
  There are infinitely many primes. & \cmark & \cmark \\
  Any function $\NN \to \NN$ is the zero function or not. & \cmark{} (trivially so) & \xmark \\
  Any function $\NN \to \NN$ is computable by a Turing machine. & \xmark & \cmark{} (trivially so) \\
  Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
  Heyting arithmetic is categorical. & \xmark & \cmark \\
  \bottomrule
\end{tabular}

\subsection{``Any natural number is prime or not.''} Even without knowing what
a prime number is, one can safely judge this statement to be true in
the standard topos, since it is just an instance of the law of excluded middle.

@@

\subsection{``There are infinitely many primes.''} A first-order formalization
of this statement is ``for any natural number~$n$, there is a prime
number~$p$ which is greater than~$n$'', and is known to be true in the standard
topos by any of the many proofs of this fact.

@@

\subsection{``Any function~$\NN \to \NN$ is the zero function or not.''} More
formally, the statement is
\[ \forall f \? \NN^\NN\_
  \bigl((\forall n \? \NN\_ f(n) = 0) \vee
  \neg
  (\forall n \? \NN\_ f(n) = 0)\bigr). \]
By the law of excluded middle, this statement is trivially true in the standard
topos.

@@

\subsection{``Any function~$\NN \to \NN$ is computable by a Turing machine.''}
The preceding examples give the impression that what is true in the
\effective topos is simply a subset of what is true in the standard topos. The
example of this subsection shows that the relation between the two toposes is more nuanced.

@@

\subsection{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
In the standard topos, this statement is plainly false, with the signum and Heaviside functions
being prominent counterexamples. In the \effective topos, this statement is
true. A rigorous proof is not entirely straightforward~\cite{XXX}, but an intuitive
explanation is as follows.

@@

* XXX Cauchy vs. Dedekind numbers (physical quantities, ...)
* XXX Syntactical vs. semantical interpretation
* XXX phone call analogy
* XXX detailed explanation of the pretty picture; analogy with "inner models"
  of set theory
* XXX quick overview of the several aspects of toposes (maybe at the end,
  as an outlook?)
* XXX allowing a switch of perspective
* XXX applications in mathematical practice
* XXX int. logic as common denominator
* XXX arbitrariness of the "standard axioms"
* XXX what giving up classical logic actually amounts to in practice
* XXX importance of having an adapted language
* XXX SDG (infinitesimals are well-studied in philosophy of mathematics)
* XXX quick remark on the internal language being based on types instead of sets
* XXX Source https://plato.stanford.edu/entries/category-theory/ for relevant
  literatur
* XXX the internal language as a generalized modal operator

* XXX use the phrase "very rich and very deep" (or similarly) about set theory
* XXX Sven(?)
* XXX semantic view vs. syntactic view
* XXX Kripke models
* XXX more pointers to topos literature
* XXX Reference for the effective topos
* XXX Reference idempotency
* XXX A topos is more than its semantics
* XXX Reference: Moerdijk
* XXX Check references everywhere (give good pointers)
* XXX Format rules in a better way
* XXX Format examples in Eff in a better way
* XXX toposes are not small (Figure 1)
* XXX hypercomputation or supercomputation?

>>>>

\begin{tabular}{lll}
  \toprule
  Statement & in $\Set$ & in $\Eff$ \\
  \midrule
  Any natural number is prime or not prime. & \cmark{} (trivially so) & \cmark \\
  There are infinitely many primes. & \cmark & \cmark \\
  Any function $\NN \to \NN$ is the zero function or not. & \cmark{} (trivially so) & \xmark \\
  Any function $\NN \to \NN$ is computable by a Turing machine. & \xmark & \cmark{} (trivially so) \\
  Any function $\RR \to \RR$ is continuous. & \xmark & \cmark \\
  Heyting arithmetic is categorical. & \xmark & \cmark \\
  \bottomrule
\end{tabular}
% XXX formatting

\newcommand{\dotparagraph}[1]{\noindent\textbf{#1}}

\bigskip
\dotparagraph{``Any natural number is prime or not.''} Even without knowing what
a prime number is, one can safely judge this statement to be true in
the standard topos, since it is just an instance of the law of excluded middle.

@@

\bigskip
\dotparagraph{``There are infinitely many primes.''} A first-order formalization
of this statement is ``for any natural number~$n$, there is a prime
number~$p$ which is greater than~$n$'', and is known to be true in the standard
topos by any of the many proofs of this fact.

@@

\bigskip
\dotparagraph{``Any function~$\NN \to \NN$ is the zero function or not.''} More
formally, the statement is
\[ \forall f \? \NN^\NN\_
  \bigl((\forall n \? \NN\_ f(n) = 0) \vee
  \neg
  (\forall n \? \NN\_ f(n) = 0)\bigr). \]
By the law of excluded middle, this statement is trivially true in the standard
topos.

@@

\bigskip
\dotparagraph{``Any function~$\NN \to \NN$ is computable by a Turing machine.''}
The preceding examples give the impression that what is true in the
\effective topos is simply a subset of what is true in the standard topos. The
example of this subsection shows that the relation between the two toposes is more nuanced.

@@

\bigskip
\dotparagraph{``Any function~$\RR \to \RR$ is continuous.''}\label{sect:eff-continuous}
In the standard topos, this statement is plainly false, with the signum and Heaviside functions
being prominent counterexamples. In the \effective topos, this statement is
true. A rigorous proof is not entirely straightforward~\cite{XXX}, but an intuitive
explanation is as follows.

@@

* XXX Syntactical vs. semantical interpretation
* XXX phone call analogy
* XXX quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
* XXX allowing a switch of perspective
* XXX applications in mathematical practice
* XXX int. logic as common denominator
* XXX arbitrariness of the "standard axioms"
* XXX importance of having an adapted language
* XXX quick remark on the internal language being based on types instead of sets
* XXX Source https://plato.stanford.edu/entries/category-theory/ for relevant literature
* XXX the internal language as a generalized modal operator
* XXX use the phrase "very rich and very deep" (or similarly) about set theory
* XXX Sven(?)
* XXX semantic view vs. syntactic view
* XXX Kripke models
* XXX more pointers to topos literature
* XXX Reference for the effective topos
* XXX Reference idempotency
* XXX A topos is more than its semantics
* XXX Reference: Moerdijk
* XXX Check references everywhere (give good pointers)
* XXX Format rules in a better way
* XXX toposes are not small (Figure 1)
* XXX hypercomputation or supercomputation?
